Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-16 doc/mc-paper.tex
Most of the technical stuff is done

% Lean code:
\usepackage{listings}
%\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
\definecolor{keywordcolor}{rgb}{0.0, 0.1, 0.6}   % blue
\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
\def\lstlanguagefiles{lstlean.tex}
% set default language
\lstset{language=lean, xleftmargin=1em}
\lstset{backgroundcolor=\color{white}}

@@

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\rep\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step maintaining logical equivalence to
the original formula.  More precisely, the status of the proof at any step will be represented by
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables}.  Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
we want to maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

@@

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic.
It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_p$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

Figure~\ref{fig:chain} illustrates one method of generating a CRAT
file from an input formula in conjunctive normal form.  Our tool chain
is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
but it would also work for other tools that use a top-down search
strategy to generate a d-DNNF representation.  Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except for the degenerate cases
where the formula is unsatisfiable or a tautology, we can also assume
that the POG does not contain any constant nodes.
In addition, negation operators will
only be applied to variables, and so we can view the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

For input formula $\inputformula$ and its translation into a POG with root literal
$r$, the most challenging part of the proof is to show that
$\modelset(\inputformula) \subseteq \modelset(P)$, i.e., that any total
assignment $\alpha$ that is a model of $\inputformula$ will yield $\assign(r) =
1$, for root literal $r$.
This part of the proof consists of a series of clause assertions
leading to one adding $r$ as a unit clause.

We describe this part of the proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target clause}
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating
that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add the unit clause $r$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

Let $\psi$ be a subset of the clauses in $\inputformula$.
For partial assignment
$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
obtained from $\psi$ by: 1) eliminating any
clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
2) for the remaining clauses eliminating those literals $\lit$ for
which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
In doing these simplifications, we also track the {\em provenance}
of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
those clauses $C' \in \psi$, such that
$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
We then extend the definition of $\prov$ to any simplified formula
$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CRAT file.
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
\end{enumerate}
  We then combine as hints the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A))
to generate the target clause for node $\nodeu$.
\end{enumerate}
Observe that most of these steps involve a polynomial number of
operations per recursive call.  The only step that could have
exponential complexity is when a SAT solver must be invoked to validate a
literal.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root literal $r$ requires showing that
$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\psi$.  Deleting $C$ requires proving that
$\modelset(\psi) = \modelset(\psi \lor C)$, i.e., that deleting $C$
does not make the formula easier to satisfy.

@@

The performance of forward implication, both in its execution time and
the size of the proof generated by two optimizations described here.
A key feature of these optimizations is that they do not require any changes to the
proof framework---they build on the power of extended resolution to
create new logical structures within the proof.

@@

A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
considered each literal $\lit \in L$ separately,
calling a SAT solver
on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ could not be validated with BCP\@.

Literal grouping allows all literals to be validated with a single call to the SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  This can be done by contradiction using the SAT solver.
In particular, looking at the defining clauses for $\nodev$ (Table
\ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
the first defining clause to simplify to
$C = \obar{\lit}_1 \lor  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
  clauses to become tautologies.  The SAT solver can be invoked on formula
  $\simplify{\psi}{\passign} \cup \{ C \}$ to show that is unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
  input clauses and the defining clauses for $\nodev$.

@@

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.  We can exploit the property
that a shared subgraph in the graph generated by a top-down d-DNNF
compiler arises when the program detects that the simplified set of
clauses arising from one partial assignment to the literals matches that
of a previous partial assignment.  When this d-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and also check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The forward implication proof can exploit this property of shared
subgraphs by constructing and proving a {\em lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma$ must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.  The net effect is to allow $\validate$ to
exploit the graph structure of the POG, invoking $\validate$ on each
node at most once.

>>>>

% Lean code:
\usepackage{listings}
\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red
\def\lstlanguagefiles{lstlean.tex}
% set default language
\lstset{language=lean, xleftmargin=1em}
\lstset{backgroundcolor=\color{white}}

@@

\begin{definition}[Ring Evaluation Problem]
  For commutative ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $\mulident-w(x)$.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
For any labeling $w$, ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\rep\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\rep\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step maintaining logical equivalence to
the original formula.  More precisely, the status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables}.  Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$, 
we want to maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

@@

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic.
It is a way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

Figure~\ref{fig:chain} illustrates one method of generating a CRAT
file from an input formula in conjunctive normal form.  Our tool chain
is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
but it would also work for other tools that use a top-down search
strategy to generate a d-DNNF representation.  Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except for the degenerate cases
where the formula is unsatisfiable or a tautology, we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation operators will
only be applied to variables, and so we can view the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root literal $r$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(P)$, i.e.,
that any total assignment $\alpha$ that is a model of $\inputformula$
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$r$ as a unit clause.  We have devised two methods of generating this
proof.  The {\em monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The {\em structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It too may require calls to a proof-generating SAT
solver.

As notation,
let $\psi$ be a subset of the clauses in $\inputformula$.
For partial assignment
$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
obtained from $\psi$ by: 1) eliminating any
clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
2) for the remaining clauses eliminating those literals $\lit$ for
which $\passign(\lit) = 0$, and 3) eliminating any duplicate clauses.
In doing these simplifications, we also track the {\em provenance}
of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov(C, \psi)$ denote
those clauses $C' \in \psi$, such that
$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
We then extend the definition of $\prov$ to any simplified formula
$\gamma$ as $\prov(\gamma, \psi) = \bigcup_{C \in \gamma} \prov(C, \psi)$.

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG, given by its set of defining clauses $\pogformula$.
Furthermore, it can
negate the POG formula by simply negating its root literal.  Forward implication
will hold
when $\inputformula \imply (r \land \pogformula)$.  Equivalently,
the formula $\inputformula \land \simplify{\pogformula}{\{\obar{r}\}}$
should be unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file, invokes a proof-generating solver, and
imports the proof steps as assertion steps in the CRAT file.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

We describe the structural approach to proof generation as a recursive procedure
$\validate(\nodeu, \passign, \psi)$ taking as arguments POG
node $\nodeu$, partial assignment
$\passign$, and a set of clauses $\psi \subseteq \inputformula$.
The procedure adds a number of clauses to the proof, culminating with
the addition of the {\em target clause}
$u \lor \bigvee_{\lit \in \passign} \obar{\lit}$, indicating 
that $\left(\bigwedge_{\lit \in \passign} \lit\right) \imply u$, i.e.,
that any total
assignment $\assign$ such that $\passign \subseteq \assign$
will assign $\assign(u) = 1$.
The top-level call has $\nodeu = \noder$, $\passign = \emptyset$, and $\psi = \inputformula$.
The result will therefore be to add the unit clause $r$ to the proof.
Here we present a correct, but somewhat inefficient formulation of
$\validate$.  We then refine it with some optimizations.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \lit'$, such that any $C' \in \prov(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ must have $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CRAT file.
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \bigcup_{\lit \in L} \lit, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \bigcup_{\lit \in L} \lit, \psi_i\right)$.
\end{enumerate}
  We then combine as hints the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A))
to generate the target clause for node $\nodeu$.
\end{enumerate}
Observe that most of these steps involve a polynomial number of
operations per recursive call.  The only one that could have
exponential complexity is when a SAT solver must be invoked to validate a
literal.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root literal $r$ requires showing that
$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\phi$.  Deleting $C$ requires proving that
$\phi \imply C$.

@@

The performance of forward implication, both in its execution time and
the size of the proof generated can be improved by two optimizations described here.
A key feature is that they do not require any changes to the
proof framework---they build on the power of extended resolution to
create new logical structures within the proof.

@@

A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
considers each literal $\lit \in L$ separately,
calling a SAT solver
on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ cannot be validated with BCP\@.

Literal grouping allows all literals to be validated with a single call to the SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof.
Looking at the defining clauses for $\nodev$ (Table
\ref{tab:defining}(A)), we can see that assigning 0 to $v$ will cause
the first defining clause to simplify to
$C = \obar{\lit}_1 \lor  \obar{\lit}_2 \lor \cdots \lor \obar{\lit}_m$, and for the remaining
  clauses to become tautologies.  The SAT solver can be invoked on formula
  $\simplify{\psi}{\passign} \cup \{ C \}$ to show that is unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
  input clauses and the defining clauses for $\nodev$.

@@

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.  
These shared subgraphs arise when the knowledge compiler employs {\em clause caching}
detecting that the simplified set of
clauses arising from one partial assignment to the literals matches that
of a previous partial assignment~\cite{darwiche:aaai:2002}.
When this d-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and also check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The proof generator can exploit the sharing of subgraphs
by constructing and proving a {\em lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma$ must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.  The net effect is that generator will recurse during a call to $\validate$ only once for each node in the POG\@.
